### 2016-05-16 Paper/paper.tex
full draft of intro and methods

\section{Introduction}
\indent ``Lagrangian Time Series Models for Ocean Surface Drifter Trajectories," by Adam M. Sykulski, Sofia C. Olhede, Jonathan M. Lilly, and Eric Danioux presents a multi-component spectral model for analysis of data transmitted by ocean surface drifters. Drifters are free-floating  instruments that transmit their location at regular time intervals,  creating a \textbf{\it{Langrangian time series}}, or a sequence of spatial locations over time. Oceanographers use these time series to increase understanding of  ocean circulation patterns. Published in January 2016 in the \textbf{\it{Journal of the Royal Statistical Society Series C}}, this paper develops a model that can isolate a scientific quantity of interest,  and, in so doing, introduces several new statistical techniques.\\
\indent The motion of parcels of water moving over changing latitudes is known to have a rotational component. Because the Earth has  different diameter at different latitudes, the speed of objects moving with Earth's rotation varies across latitudes.  This induces a phenomena known as the Coriolis Effect, whereby very fast moving objects or objects being observed over long periods, such as drifters, end up moving at speeds that are different than the Earth beneath them as they change latitude. Viewed from a stationary reference frame, this creates circular motions, or inertial oscillations. These motions have known frequencies, referred to as inertial frequencies. Oceanographers  are interested in detecting deviations from these frequencies, as they are thought to indicate eddies, or persistent circular wave patterns \citep{Kunze1985}. However, the ocean has a constant turbulent background flow which makes identifying eddies in real data difficult \citep{ Elipot2010}. Previous methods to model drifters are BLAH The methods in this paper overcome this challenge and successfully find shifts in the inertial frequency with a spectral time series model. \\
\indent In achieving this result, several statistical advances are made.  The authors introduce an additive spectral domain model that has components corresponding to both the turbulent background and the inertial oscillations. For the turbulent background model, the Mat\'{e}rn covariance commonly used in spatial statistics \citep{Gneiting2012} is extended to the time-series context. This allows for a more flexible model for background behavior than previously obtained.  Inertial oscillations are modeled with a complex Ornstein-Uhlenbeck (OU) \citep{Arato1962, Jeffreys1968}, which provides a stochastic equivalent to an accepted set of coupled differential equations describing inertial oscillations. Further statistical innovation is employed in fitting the model with a variation on standard Whittle likelihood that reduces bias. Time-varying parameter estimates are also introduced to allow for non-stationarity.  \\
\indent The remainder of this review proceeds as follows. After a succinct review of spectral analysis, Section 2 discusses the model for the drifters and the methods used to fit it. Section 3 presents the paper's results including analysis of a long time series with time-varying parameters and an application from a simulated model. Section 4 concludes the review with a discussion of the value and limitations of the approach taken by \citet{Sykulski2016}. 
\section{Methods}

@@

\begin{figure}[hb]
	\label{fig: fmbMat}
  \centering
    \includegraphics[width=.5\textwidth]{fig3.pdf}
        \caption{Simulated velocities obtained using FBM (blue) and Mat\'{e}rn (red)  processes. Top Left: The eastward component the velocity; Top Right: The eastward and northward components of the vecocity; Bottom: Periodograms on the decibel scale}
\end{figure}

@@

\subsection{Fitting the Model}
\subsubsection{Blurred Whittle likelihood}
A new technique introduced in  \citet{Sykulski2013}  is used to fit the model. Referred to as \textbf{\it{blurred Whittle likelihood}}, this new likelihood builds on the well-known Whittle likelihood, a frequency-based approximation  to the time-domain likelihood that reduces computational cost.  Whittle likelihood is designed to approximate the likelihood for a univariate Gaussian process, $Z$,  in the time domain that has the likelihood
\begin{align*}
l_{t}(\theta) = - \frac{1}{2} log |C_{Z}(\theta) | - \frac{1}{2} Z^{T} C_{Z}^{-1}(\theta)Z
\end{align*}
where $\theta$ are the model parameters for a particular choice of covariance structure  and $t$ denotes the time domain  \citep{Sykulski2013}.  It is known that  the periodogram, given in Equation ~\ref{eq: perio}, satisfies
\begin{equation}
|J_{Z}(\omega)| \overset{d}{\rightarrow} S_{Z}(\omega) \chi^{2}_{2}/2
\end{equation}
under certain regularity conditions CITE. \citet{Whittle1953} combined this result with Fourier approximations to the true covariance structure and features of Toeplitz matrices to obtain the approximate likelihood
\begin{align}
l_{\omega}(\theta) &= - \sum_{\omega \in \Omega} \left[ J_{Z}^{H}(\omega) \tilde{S}_{Z}^{-1}(\omega; \theta) J_{Z}(\omega) + \log \{S(\omega; \theta) \} \right]\\
\label{eq: bwl}
&= -\sum_{\omega \in \Omega} \left[ \frac{\hat{S}_{Z}(\omega)}{S(\omega;\theta)}  + \log  \{ S(\omega; \theta) \}\right]
\end{align}
 where the subscript $w$ indicates that this is a Whittle likelihood and $H$ denotes a Hermitian matrix.  However, this likelihood relies on the periodogram and so is affected by aliasing and leakage as discussed in Section ~\ref{sec: specAnalysis}.

To improve on these problems, \citep{Sykulski2013} propose replacing the theoretical spectrum, $S_{Z}(\omega; \theta)$ with the expected value of the spectrum, $\overline{S}_{Z}(\omega, \theta)$\begin{equation}
\overline{S}(\omega; \theta) = \Delta \sum_{\tau = - (N - 1)}^{N-1} \left(1 - \frac{|\tau|}{N} \right)s_{\tau}(\theta) exp ( - i \omega \tau \Delta)
\end{equation}
in Equation ~\ref{eq: bwl}. This form of the expected spectrum can be shown to be equivalent to convolving the theoretical spectrum with the Fe\'{j}er kernel 
\begin{equation}
\mathcal{F}(\cdot) = \frac{\Delta}{2\pi N} \frac{sin^{2}(N \omega \Delta/2)}{sin^{2}(\omega \Delta /2)} 
\end{equation} \citep{Sykulski2013}. This improves the estimators obtained through maximizing the likelihood, because the expected value of the periodogram $\mathbb{E}[|J_{Z}(\omega)^{2}]$ is blurred in the same way as the observed periodogram; thereby eliminating effects due to sampling at discrete intervals. This leads to the final likelihood
\begin{align}
l_{b}(\theta) = - \sum_{\omega \in \Omega} \left[\frac{\hat{S}_{Z}(\omega)}{\overline{S}(\omega; \theta)} + \log \{ \overline{S} (\omega; \theta) \}\right]
\end{align}
which is maximized by standard optimization techniques.

@@

\subsubsection{Significance tests for simpler models}
Finally, \citet{Sykulski2016} propose that a simple likelihood ratio test can be used at each time period to test whether or not a shift in the inertial oscillation has occurred. To do so, the authors suggest fitting a simpler model where the the parameter $\omega_{0}$ is fixed to be the inertial frequency $f$. Then this simpler model can be compared to the full six parameter model via a likelihood ratio test with test statistic
\begin{equation}
\label{eq: LRT}
R(t) = 2[l_{b} \{\hat{\theta}_{1}(t) \} - l_{b}\{\hat{\theta}_{0}(t) \} ]
\end{equation}
where $\hat{\theta}_{0}(t)$ is the fitted parameter vector for the simpler, or null model, and $\hat{\theta}_{1}(t)$ is the fitted parameter vector for the full model. The authors claim $R(t)$ then follow a Chi-Squared distribution with 1 degree of freedom.

>>>>

\section{Introduction}
\indent ``Lagrangian Time Series Models for Ocean Surface Drifter Trajectories," by Adam M. Sykulski, Sofia C. Olhede, Jonathan M. Lilly, and Eric Danioux presents a multi-component spectral model for analysis of data transmitted by ocean surface drifters. Drifters are free-floating  instruments that transmit their location at regular time intervals,  creating a \textbf{\it{Langrangian time series}}, or a sequence of spatial locations over time. Oceanographers use these time series to increase understanding of  ocean circulation patterns. Published in January 2016 in the \textbf{\it{Journal of the Royal Statistical Society Series C}}, this paper develops a model that can isolate a scientific quantity of interest,  and, in so doing, introduces several new statistical techniques.\\
\indent The motion of parcels of water moving over changing latitudes is known to have a rotational component. Because the Earth has  different diameter at different latitudes, the speed of objects moving with Earth's rotation varies across latitudes.  This induces a phenomena known as the Coriolis Effect, whereby very fast moving objects or objects being observed over long periods, such as drifters, end up moving at speeds that are different than the Earth beneath them as they change latitude. Viewed from a stationary reference frame, this creates circular motions, or inertial oscillations. These motions have known frequencies, referred to as inertial frequencies. Oceanographers  are interested in detecting deviations from these frequencies, as they are thought to indicate eddies, or persistent circular wave patterns \citep{Kunze1985}. However, the ocean has a constant turbulent background flow which makes identifying eddies in real data difficult \citep{ Elipot2010}. The methods in this paper overcome this challenge and successfully find shifts in the inertial frequency with a spectral time series model. \\
\indent In achieving this result, several statistical advances are made.  The authors introduce an additive spectral domain model that has components corresponding to both the turbulent background and the inertial oscillations. For the turbulent background model, the Mat\'{e}rn covariance commonly used in spatial statistics \citep{Gneiting2012} is extended to the time-series context. This allows for a more flexible model for background behavior than previously obtained.  Inertial oscillations are modeled with a complex Ornstein-Uhlenbeck (OU) \citep{Arato1962, Jeffreys1968}, which provides a stochastic equivalent to an accepted set of coupled differential equations describing inertial oscillations. Further statistical innovation is employed in fitting the model with a variation on standard Whittle likelihood that reduces bias. Time-varying parameter estimates are also introduced to allow for non-stationarity.  \\
\indent The remainder of this review proceeds as follows. After a succinct review of spectral analysis, Section 2 discusses the model for the drifters and the methods used to fit it. Section 3 presents the paper's results including analysis of a long time series with time-varying parameters and an application from a simulated model. Section 4 concludes the review with a discussion of the value and limitations of the approach taken by \citet{Sykulski2016}. 
\section{Methods}

@@

\begin{figure}[hb]
	\label{fig: fmbMat}
  \centering
    \includegraphics[width=.5\textwidth]{fig3.pdf}
        \caption{Simulated velocities obtained using FBM (blue) and Mat\'{e}rn (red)  processes. Top Left: The eastward component the velocities. Top Right: The eastward and northward components of the velocities. Bottom: Periodograms of both processes on the decibel scale. }
\end{figure}

@@

\subsection{Fitting the Model}
\subsubsection{Blurred Whittle likelihood}
A new technique introduced in  \citet{Sykulski2013}  is used to fit the model. Referred to as \textbf{\it{blurred Whittle likelihood}}, this new likelihood builds on the well-known Whittle likelihood, a frequency-based approximation  to the time-domain likelihood that reduces computational cost.  Whittle likelihood is designed to approximate the likelihood for a univariate Gaussian process, $Z$,  in the time domain that has the likelihood
\begin{align*}
l_{t}(\theta) = - \frac{1}{2} log |C_{Z}(\theta) | - \frac{1}{2} Z^{T} C_{Z}^{-1}(\theta)Z
\end{align*}
where $\theta$ are the model parameters for a particular choice of covariance structure  and $t$ denotes the time domain  \citep{Sykulski2013}.  It is known that  the periodogram, given in Equation ~\ref{eq: perio}, satisfies
\begin{equation}
|J_{Z}(\omega)| \overset{d}{\rightarrow} S_{Z}(\omega) \chi^{2}_{2}/2
\end{equation}
under certain regularity conditions \citep{Contreras2006}. \citet{Whittle1953} combined this result with Fourier approximations to the true covariance structure and features of Toeplitz matrices to obtain the approximate likelihood
\begin{align}
l_{\omega}(\theta) &= - \sum_{\omega \in \Omega} \left[ J_{Z}^{H}(\omega) \tilde{S}_{Z}^{-1}(\omega; \theta) J_{Z}(\omega) + \log \{S(\omega; \theta) \} \right]\\
\label{eq: bwl}
&= -\sum_{\omega \in \Omega} \left[ \frac{\hat{S}_{Z}(\omega)}{S(\omega;\theta)}  + \log  \{ S(\omega; \theta) \}\right]
\end{align}
 where the subscript $w$ indicates that this is a Whittle likelihood and $H$ denotes a Hermitian matrix.  This likelihood can be evaluated in $\mathcal{O}(N log(N))$ where $N$ is the length of the time series which is a notable increase in speed compared to the $\mathcal{O}(N^{3}$  required for time-domain likelihood  \citep{Sykulski2016}. Nevertheless, this likelihood relies on the periodogram and so is affected by aliasing and leakage as discussed in Section ~\ref{sec: specAnalysis}.

To improve on these problems, \citep{Sykulski2013} propose replacing the theoretical spectrum, $S_{Z}(\omega; \theta)$ with the expected value of the spectrum, $\overline{S}_{Z}(\omega, \theta)$\begin{equation}
\overline{S}(\omega; \theta) = \Delta \sum_{\tau = - (N - 1)}^{N-1} \left(1 - \frac{|\tau|}{N} \right)s_{\tau}(\theta) exp ( - i \omega \tau \Delta)
\end{equation}
in Equation ~\ref{eq: bwl}. This form of the expected spectrum can be shown to be equivalent to convolving the theoretical spectrum with the Fe\'{j}er kernel 
\begin{equation}
\mathcal{F}(\cdot) = \frac{\Delta}{2\pi N} \frac{sin^{2}(N \omega \Delta/2)}{sin^{2}(\omega \Delta /2)} 
\end{equation} \citep{Sykulski2013}. This improves the estimators obtained through maximizing the likelihood, because the expected value of the periodogram $\mathbb{E}[|J_{Z}(\omega)^{2}]$ is blurred in the same way as the observed periodogram; thereby eliminating effects due to sampling at discrete intervals. This leads to the final likelihood
\begin{align}
l_{b}(\theta) = - \sum_{\omega \in \Omega} \left[\frac{\hat{S}_{Z}(\omega)}{\overline{S}(\omega; \theta)} + \log \{ \overline{S} (\omega; \theta) \}\right]
\end{align}
which is maximized by standard optimization techniques. Using asymptotic properties of the Whittle likelihood, we also have 
\begin{equation}
\label{eq: fisher}
\sqrt{N}(\hat{\theta} - \theta) \overset{d}{\rightarrow} \text{ Normal($0, \mathcal{F}^{-1})$ }.
\end{equation}
where $\mathcal{F}$ is the Fisher information. Thus, uncertainty estimates of the parameters can easily be obtained using numerical calculations of the Hessian.

@@

\subsubsection{Significance tests for simpler models}
Finally, \citet{Sykulski2016} propose that a simple likelihood ratio test can be used at each time period to test whether or not a shift in the inertial oscillation has occurred. To do so, the authors suggest fitting a simpler model where the the parameter $\omega_{0}$ is fixed to be the inertial frequency $f$. Then this simpler model can be compared to the full six parameter model via a likelihood ratio test with test statistic
\begin{equation}
\label{eq: LRT}
R(t) = 2[l_{b} \{\hat{\theta}_{1}(t) \} - l_{b}\{\hat{\theta}_{0}(t) \} ]
\end{equation}
where $\hat{\theta}_{0}(t)$ is the fitted parameter vector for the simpler, or null model, and $\hat{\theta}_{1}(t)$ is the fitted parameter vector for the full model. The authors claim $R(t)$ then follow a Chi-squared distribution with 1 degree of freedom.
